### 2020-08-15 manuscript/manuscript.rmd
resolved some comments and other minor edits

There has been considerable debate about the placement of the root in the animal tree of life. This question has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system. Some phylogenetic analyses support a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others find support for a split between comb jellies and all other animals (Ctenophora-sister). [XXXIt would be helpful if you also stated why this is a problem: Antonis] These analyses differed in many respects, including genes considered, species considered, molecular evolution models, and software. Here we systematically explore this question under consistent conditions by synthesizing data and results from 15 previous phylogenomic studies and performing new standardized analyses. Published studies find support for Porifera-sister in some analyses with reduced outgroup (non-animal) sampling and complex site heterogeneous models, and Ctenophora-sister under other conditions. Our new analyses are consistent with this. Though this had been interpreted by some to suggest Ctenophora-sister is an artifact of amino-acid composition bias, our analyses instead indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained overfit models with a very high number of substitutional categories. [XXXTo me this part is a diversion – I would state your key inference (“our analyses indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained overfit models with a very high number of site categories”) upfront. You can then explain how your results fit / shed light on previous interpretations of certain analyses.]Our analyses help reconcile variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for development of phylogenomic methods.

@@

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Figs. 1B, 1E) was widely accepted though rarely tested. By contrast, there has long been uncertainty about the placement of Ctenophora in the animal tree of life [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative (Figs. 1A, 1D). Since then, many more relevant phylogenomic studies have been published, with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Extended Data Table 1). The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result (Fig. 1). But the extensive technical variation has also made it difficult to synthesize these results to understand the underlying causes of this sensitivity. This methodological variance is often confounded, for example when a particular model is only available in one software package.

@@

The models of sequence evolution in the studies considered here differ according to two primary components: the exchangeability matrix $E$ and amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $E$ describes the relative rates at which one amino acid changes to another. The studies considered here use exchangeabilities that are the same between all amino acids (Poisson, also referred to as F81),or different; if different, the exchangeabilities can either be fixed based on previously empirically estimated rates (WAG or LG), or independently estimated from the data (GTR). The analyses considered here have site-homogeneous exchangeability models (site-homogeneous model), with the same matrix used for all sites in all genes. Or for data-partitioning analyses, some gene partitions use different site-homogeneous models. The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others. The analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites in a partition, or add parameters that allow heterogeneous equilibrium frequencies that differ across sites. Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention. While studies often discuss CAT and WAG models as if they are exclusive, we note that these particular terms apply to non-exclusive model components – CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, here we always specify the exchangeability matrix first (*e.g.* GTR), followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.* CAT). If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone. Gamma-rate heterogeneity, a scalar that accommodates total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.

@@

The observation that support across genes for Ctenophora-sister is not sensitive to outgroup sampling under site homogenous models, but is sensitive under site heterogeneous models, indicates that XX.

@@

Overcoming the challenges associated with this difficult and important question goes beyond models. In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic context [@King:2017ie]. In particular, there are very few chromosome-level genome assemblies outside of Bilateria. There will be important advances in genome sequencing in coming years that will have a major impact on this problem. Instead of continuing to use previous matrices that has limited taxa sampling (*e.g.* Chang2015 and Whelan2015), without analyses with greater taxon sampling, confound conclusions as greater taxon sampling should result in better tree inference. [XX Nathan: But, I think there needs to be a call for generating datasets with the most available non-bilaterian taxon sampling. Continuing to use datasets from Chang and Whelan, without analyses with greater taxon sampling, confound conclusions as greater taxon sampling should result in better tree inference.] This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination. We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.

@@

**Extended Data Fig. 5.** (A) The two alternative hypotheses for deep animal relationships considered here. Relationships that are not part of these hypotheses are shown as unresolved polytomies. (B) Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses. These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that I can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 - random-03). Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point). For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding. This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.

@@

**Extended Data Fig. 6.** The subset of eight GTR+CAT analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions. These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane. The eight analyses are for two datasets (Chang and Whelan) and four coding schemes. The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6. Cells are color coded as in Supplemental Fig XXRecoidng_summary. Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister. The 20-state and 6-state points on the plots in Fig XXRecoidng_summary of the main text correspond to the 20 and SR6 Whelan cells shown here. The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy). The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the posterior predictive scores, and showed the poorest convergence according to Maxdiff.

@@

# Supplementary Information

### S.1 Models of molecular evolution
Models of molecular evolution are assembled by selecting different options for all these different components. The models that are applied in practice are heavily influenced by computational costs, as well as convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. studies often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. One could, though, run a WAG+CAT model, although the computation time is much longer than Poisson+CAT model. To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.* CAT).  Moreover, the Gamma-rate heterogeinity (G4) is used in almost every analysis conducted here. If there are no modifiers, then it is implied that site homogeneous models are used.

- Recoding method:  Amino acids are recoded into six groups based on function to account for both compositional heterogeneity and substitution saturation. Several recoding strategies have been proposed, including Dayhoff 6-state recoding, S&R 6-state recoding, KGB 6-state recoding based on exchangeability matrix of models.

### S.2 Details of published analyses

@@

>>>>

There has been considerable debate about the placement of the root in the animal tree of life, which has emerged as one of the most difficult problems in animal phylogenetics in part because this is the oldest part of the animal tree. This question has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system. Some phylogenetic analyses support a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others find support for a split between comb jellies and all other animals (Ctenophora-sister). These analyses differed in many respects, including genes considered, species considered, molecular evolution models, and software. Here we systematically explore this question under consistent conditions by synthesizing data and results from 15 previous phylogenomic studies and performing new standardized analyses. Published studies find support for Porifera-sister in some analyses with reduced outgroup (non-animal) sampling and complex site heterogeneous models, and Ctenophora-sister under other conditions. Our new analyses are consistent with this. Our analyses indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained models with a very high number of site categories. Our analyses help reconcile variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for development of phylogenomic methods.

@@

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Figs. 1B, 1E) was widely accepted though rarely tested. By contrast, there has long been uncertainty about the placement of Ctenophora in the animal tree of life [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative (Figs. 1A, 1D). Since then, many more relevant phylogenomic studies have been published, with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Extended Data Table 1, Supplementary Information section S1). The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result (Fig. 1). But the extensive technical variation has also made it difficult to synthesize these results to understand the underlying causes of this sensitivity. This methodological variance is often confounded, for example when a particular model is only available in one software package.

@@

The models of sequence evolution in the studies considered here differ according to two primary components: the exchangeability matrix $E$ and amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $E$ describes the relative rates at which one amino acid changes to another. The studies considered here use exchangeabilities that are the same between all amino acids (Poisson, also referred to as F81),or different; if different, the exchangeabilities can either be fixed based on previously empirically estimated rates (WAG or LG), or independently estimated from the data (GTR). The analyses considered here have site-homogeneous exchangeability models (site-homogeneous model), with the same matrix used for all sites. The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others. The published analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites in a partition, or add parameters that allow heterogeneous equilibrium frequencies that differ across sites. Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention. While studies often discuss CAT and WAG models as if they are exclusive, we note that these particular terms apply to non-exclusive model components – CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, here we always specify the exchangeability matrix first (*e.g.* GTR), followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.* CAT). If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone. Gamma-rate heterogeneity, a scalar that accommodates total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation. Some analyses partition the data by genes, and use different models for each gene.

@@

@@

Overcoming the challenges associated with this difficult and important question goes beyond models. In particular, there are many organisms that are highly relevant to this problem, in particular better sampling of outgroups, ctenophores, and sponges, for which no genome or transcriptome data are available [@King:2017ie]. In particular, there are very few chromosome-level genome assemblies for animals outside of Bilateria. There will be important advances in genome sequencing in coming years that will have a major impact on this problem. This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination. We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.

@@

**Extended Data Fig. 5.** Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses. These six analyses (XX there are only five, one is missing) are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that we can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 - random-03). Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point). For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding. This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.

@@

**Extended Data Fig. 6.** The subset of eight GTR+CAT analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions. These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane. The eight analyses are for two datasets (Chang and Whelan) and four coding schemes. The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6. Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister. The 20-state and 6-state points on the plots in Fig XXRecoidng_summary of the main text correspond to the 20 and SR6 Whelan cells shown here. The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy). The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the posterior predictive scores, and showed the poorest convergence according to Maxdiff.

@@

## Supplementary Information

### S.1 Summaries of published analyses

Narrative summaries of the studies considered here.

@@

### S.2 Models of molecular evolution
